---
title: "R"
author: "Faith"
date: "2023-04-1"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cosmo
    collapsed: true
    smooth_scroll: true
  editor_options: 
    markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE) 
# hides messages & warnings
```

In this page I will be plotting the necessary figures for the project. Functions from the file R.r will be extracted to aid in the plotting of figures. Data processing will be conducted before any figures are generated.


## Setup

### Load packages, plot setup and necessary functions

```{r echo=TRUE}
wd<- setwd(dirname(rstudioapi::getActiveDocumentContext()$path)) # set working directory

# Navigate up to parent directory
setwd("..")
wd <- getwd()

library(plyr) # data wrangling
library(dplyr) # data wrangling
library(tidyverse) # data wrangling and management
library(ggplot2) # visualization
library(vegan) # Ecological statistic tools 
library(scatterpie) # Scatterpie plot
library(smplot2) # Statistical data visualization that complements ggplot2
library(gridExtra) # Extensions to the grid system in ggplot2
library(treemapify) # Extensions to create treemaps in ggplot2
library(viridis) # Colourblind friendly map
library(rstudioapi) #Access the RStudio API 
library(scales) # Customize axis and legend labels
library(cowplot) # Supplement ggplot2
library(ggallin) #Extra geoms and scales for ggplot2

# Create base map
worldmap <- map_data ("world")# From the tidyverse package

base_world  <- ggplot()+ 
  coord_fixed() +
  xlab("") + ylab("") + 
  geom_polygon(data=worldmap, aes(x=long, y=lat, group=group), 
               fill="grey80")+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        panel.background = element_rect(fill = 'white', colour = 'white'), 
        axis.line = element_line(colour = "white"), #legend.position="none",
        axis.ticks=element_blank(), axis.text.x=element_blank(),
        axis.text.y=element_blank())

# Source for functions 
source("R/R.R")
```

## Part 0: Loading data sets and data processing
Since the data set from MetaPR2 is not updated, using the updated samples_edit.csv file I will be merging information to update the MetaPR2 dataset 

### Loading data sets
```{r echo=TRUE}

asv_sample_raw<-read_tsv(file = paste0(wd,'/DATA/METAPR2/woa_metapr2_ASVs_selected_abundance_Eukaryota.tsv')) # Change the name of your MetaPR2 file accordingly 

sample_info<- read.csv(file = paste0(wd,'/DATA/METAPR2/samples_edit.csv'))
```

### Updating MetaPR2 data
```{r echo=TRUE}
# Remove the non-mircobial eukaryotes 
asv_sample_raw<-asv_sample_raw[!(asv_sample_raw$division == "Metazoa" | asv_sample_raw$division == "Fungi"),]


# Merging information from updated samples data set to the asv data set by matching file_code

asv_sample<- asv_sample_raw %>%
  dplyr::select(-c("climate"))%>%
  merge(sample_info[,c("file_code","climate","ss_temperature","salinity_woa","coalesce_salinity")],by=c('file_code'),all.x = T)
```


### Data processing
```{r echo=TRUE}
# Reassign depth level to avoid overlap
asv_sample<-asv_sample %>%
  mutate(depth_level = case_when((depth >=  0 & depth <= 10) ~ "surface",
                                 (depth >  10 & depth <= 200 ) ~ "euphotic",
                                 (depth >  200) ~ "pelagic",
                                 TRUE ~ depth_level))
sample_info<-sample_info %>%
  mutate(depth_level = case_when((depth >=  0 & depth <= 10) ~ "surface",
                                 (depth >  10 & depth <= 200 ) ~ "euphotic",
                                 (depth >  200) ~ "pelagic",
                                 TRUE ~ depth_level))

# Correct dataset ID 385 as it was mistakenly classified as oceanic
asv_sample<- asv_sample %>%
  mutate(ecosystem = case_when((dataset_id ==  385) ~ "coastal",
                                 TRUE ~ ecosystem))

sample_info<- sample_info %>%
  mutate(ecosystem = case_when((dataset_id ==  385) ~ "coastal",
                               TRUE ~ ecosystem))
```


## Part 1: Distribution of laby (surface)
For global distribution plots, only surface communities will be plotted and time series data will be removed to avoid overlapping at one point

```{r echo=TRUE}
# Create file directory for plots from part 1
dir.create(paste0(wd,'/PLOTS/R'))
dir.create(paste0(wd,'/PLOTS/R/P1'))

# Remove unessary dataset
asv_sample_surface_nts<-asv_sample %>%
  subset(depth_level == 'surface')%>%
  subset(dataset_id!="16" & dataset_id!="18"& dataset_id!="19"& dataset_id!="20"& dataset_id!="28"& dataset_id!="29"& dataset_id!="36"& dataset_id!="49"& dataset_id!="384")
```


### Preparing Presence/absence data
```{r echo=TRUE}
laby_PAD <- asv_sample_surface_nts %>%
  group_by(file_code, latitude, longitude, class, n_reads) %>%
  dplyr::summarise(total_count = n() * n_reads, .groups = 'drop') %>% # Find the total count for each location
  dplyr::select(-n_reads) %>% group_by(file_code, latitude, longitude, class) %>%
  dplyr::summarise(total_count = sum(total_count), .groups = 'drop') %>% # Merging values of the same location
  spread(key = class, value = total_count) %>% # Convert to wide data
  dplyr::select(file_code, latitude, longitude,Labyrinthulomycetes)%>%
  mutate(P_A = case_when((Labyrinthulomycetes > 0) ~ "present",
                         is.na(Labyrinthulomycetes) ~ "absent"))%>%
  as.data.frame()
```

### Prepare data for Laby dominance map 

```{r echo=TRUE}

# Filter for class laby 
laby_asv_sample<- filter(asv_sample,class == "Labyrinthulomycetes")
laby_asv_sample_surface_nts <- filter(asv_sample_surface_nts,class == "Labyrinthulomycetes")

# Find the dominant taxon in each sample
dominant_taxon <- laby_asv_sample_surface_nts %>%
  group_by(file_code, order,n_reads_pct) %>%
  arrange(file_code, desc(n_reads_pct)) %>%
  group_by(file_code) %>%
  dplyr::slice(1) %>%
  mutate(dominant_taxon = .data[["order"]]) %>%
  dplyr::select(file_code,latitude,longitude ,dominant_taxon,n_reads_pct)

# Merge information on P/A and dominance
dominant_merge <- laby_PAD %>%
  merge(dominant_taxon[,c("file_code","dominant_taxon","n_reads_pct")], 
        by=c('file_code'),all.x = T)

```

### Plot individual components of the map

Separate plots will be created and merged together on a separate software Affinity Designer

#### Plotting main map
```{r echo=TRUE}
# Creating colour palette for the 6 orders
vir_order<- c("Amphifilida" = "#440154", "Amphitremida" = "#414487","Labyrinthulida" = "#2a788e","Labyrinthulomycetes_X" = "#22a884","Oblongichytrida" = "#7ad151","Thraustochytrida" = "#fde725")

# Create the main map
main_map<- base_world +
  geom_point(data=subset(dominant_merge, is.na(dominant_taxon)), 
             aes(x = longitude, 
                 y = latitude,
                 shape='NA'), stroke=0.8, shape=4,color = "red") +
  geom_jitter(data = dominant_merge, aes(x = longitude, 
                                         y = latitude,
                                         color=dominant_taxon,
                                         size = n_reads_pct*2),
              alpha=0.6)+
  guides(size = guide_legend(reverse=T))+
  scale_size_continuous(range  = c(0.1, 10), 
                        limits = c(0, 40), 
                        breaks = c(1, 5, 20),
                        trans=pseudolog10_trans)+
  scale_color_manual(values=vir_order)+
  labs(size="% of microbial \n eukaryotes", color="Dominant taxon")+
  theme_minimal()+
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        legend.box.background = element_rect(colour = NA))

# Add boxes to the main map indicating areas to be zoomed in
main_map_plot<- main_map +
  geom_polygon(aes(x=c(-7,-7,36,36),y=c(30,60,60,30)),
               fill = "NA", colour = "black",linewidth = 0.4)+
  geom_polygon(aes(x=c(-98,-98,-55,-55),y=c(13,48,48,13)),
               fill = "NA", colour = "black",linewidth = 0.4)+
  geom_polygon(aes(x=c(164,164,185,185),y=c(-60,-30,-30,-60)),
               fill = "NA", colour = "black",linewidth = 0.4)+
  geom_polygon(aes(x=c(-80,-80,-50,-50),y=c(-70,-60,-60,-70)),
               fill = "NA", colour = "black",linewidth = 0.4)


ggsave(filename=paste0(wd,"/PLOTS/R/P1/main_map.pdf"),plot=main_map_plot,width=40,height=19,units=c("cm"),bg="white")
```


#### Plotting zoomed in sections of map
```{r echo=TRUE}
# Using these coordinates, create small individual boxes of maps and save them individually 

# Create a dataframe to store the plots coordinates
cord_df<- data.frame(plot_num = c("A","B","C","D"),
                     x1 = c(-7,-98,164,-80),
                     x2 = c(36,-55,185,-50),
                     y1 = c(30,13,-60,-70),
                     y2 = c(60,48,-30,-60))

# Plot the zoomed in sections and save them
for (i in 1:nrow(cord_df)){
  
  small_p<- main_map+
    coord_cartesian(xlim =c(cord_df[i,"x1"], cord_df[i,"x2"]), ylim = c(cord_df[i,"y1"], cord_df[i,"y2"]))+
    theme(legend.position="none",
          panel.border = element_rect(colour = "black", fill=NA, size=0.3))
  
  ggsave(filename=paste0(wd,"/PLOTS/R/P1/small_plot",cord_df[i,"plot_num"],".pdf"),plot = small_p,dpi=300,bg="white")
}
```

## Part 2: Treemap+ scatterpie maps of each ecosystem variable
Using the function p2_plot, I will be able to plot all the individual components of the Tree map and scatterpie map required and save them. The figures will later be assembled in Affinity Designer. The function of p2_plot also gives us an output that is a list of information about the composition of different Labyrinthulomycetes orders for the different input ecosystems.

### Plot Treemap+ scatterpie maps
```{r echo=TRUE}
# Using the function, plot the different treescatterpie map according to the variables 
variable_df <- data.frame(variable=c("ecosystem","depth_level","climate"))

f_labels_combi <- list() #  New dataframe to store information of plots

for (i in variable_df[,1]){
  f_lab<- p2_plot(i)
  f_labels_combi[[i]] <- f_lab # add it to your list
}

```


### Plot depth map 
Since a map will not be meaningful in studying the vertical composition of the different Labyrinthulomycetes orders. A separate depth plot will be created to illustrate the changes in community structure in relation to depth.
```{r echo=TRUE}
# Prepare data for depth plot
laby_depth<- laby_asv_sample %>%  # for depth_level
  mutate(across(depth_level, factor, levels=c("surface","euphotic","pelagic")))%>%
  filter(ecosystem == "oceanic") %>% # Filter for laby and oceanic samples
  mutate(depth = fct_rev(cut(depth, breaks = c(0, 2.5, 5, 7.5, 10, 50, 100, 150,200, 1000,2000,3000,9000))))%>%
  drop_na(depth)%>%
  group_by(depth,order) %>%
  summarize(n_reads_pct = sum(n_reads_pct)) %>%
  group_by(depth) %>% 
  mutate(n_reads_pct = n_reads_pct/sum(n_reads_pct)*100)

# Plotting out depth plot 
depth_plot<- ggplot(laby_depth)+
  geom_col(aes (x= depth,
                y=n_reads_pct, 
                fill=order)) +
  scale_fill_viridis_d()+
  theme_minimal()+
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none")+
  scale_y_continuous(breaks = seq(0,100,20)) +
  ylab("% of reads") + xlab("Depth (m)") +
  coord_flip()

ggsave(filename=paste0(wd,"/PLOTS/R/P2/Laby_depth_plot.pdf"),plot=depth_plot,width=40,height=40,units=c("cm"),bg="white")
```

## Part 3: NMDS 
NMDs ordination plots will be used to illustrate the clustering of samples belonging to different variables. 

In this part, I will be using 2 functions, fun_norm to normalise data and nmds_plot to plot NMDs plots according to the ecosystem variables.

### Using the function, plot the NMDs plots according to the variables 
```{r echo=TRUE}
nmds_variable_df <- data.frame(variable=c("surface","depth"))

for (i in nmds_variable_df[,1]){
  nmds_plot(i,laby_asv_sample)
}

```













